---
title: Drug Arrests and Overdose Deaths in NYC
subtitle: "Evidence from NYC Data"
author:
  - name: Jordan Alfano 
    affiliation: SUNY Geneseo
    
  - name: Jake Starkey
    affiliation: SUNY Geneseo
    
date: last-modified

execute: 
  echo: true
  eval: true
  warning: false
  message: false
  fig-width: 10
  fig-height: 6

format:
  html:
    code-fold: true
    toc: true
    toc-title: "Contents"
    toc-expand: 3
    number-sections: true
    code-tools: true
    highlight-style: atom-one
---

1. Introduction 
Purpose of section
New York City has experienced a sharp rise in overdose deaths over the past decade, reflecting the broader opioid and fentanyl epidemic occurring across the United States. The crisis intensified during and after the COVID-19 pandemic as overdose deaths reached record highs across many urban communities. In response, policymakers, public health officials, and law enforcement agencies have continued to debate the effectiveness of different strategies aimed at reducing overdose deaths, particularly the role of drug-related arrests and enforcement activity.
At first glance, the relationship between drug arrests and overdose deaths appears relatively straightforward. Areas with higher levels of drug enforcement often show lower overdose death rates in simple regression models, suggesting that arrests may reduce overdose activity. However, this relationship is difficult to interpret causally because law enforcement resources are often concentrated in neighborhoods already experiencing severe drug activity. As a result, arrests may respond to overdose risk rather than directly reducing overdose deaths. This creates a reverse causality problem that complicates simple regression analysis.
This study examines the relationship between drug arrests and overdose death rates across New York City community districts using fixed effects regression analysis and neighborhood-level overdose data. The analysis combines overdose death data from the New York City Department of Health and Mental Hygiene (DOHMH), drug arrest data from the New York Police Department (NYPD), and demographic and socioeconomic data from the American Community Survey (ACS) between 2013 and 2021.
The paper also expands beyond the regression analysis by examining more recent geographic overdose patterns throughout New York City using 2023 and 2024 DOHMH neighborhood overdose data. While the regression analysis focuses on long-run relationships between arrests and overdose deaths, the later descriptive analysis is used to examine whether overdose activity remains geographically concentrated within persistent hotspot regions. Together, these approaches help evaluate whether overdose deaths are more closely associated with enforcement levels or broader neighborhood conditions.
Data preview
To examine this relationship more closely, this project uses panel data from New York City community districts between 2013 and 2021, combining overdose death data from the New York City Department of Health and Mental Hygiene, drug arrest data from the NYPD, and socioeconomic controls from the American Community Survey. Because some overdose counts were suppressed in the public health data, the analysis incorporates lower-bound, midpoint, and upper-bound overdose estimates to test the sensitivity of the results under different assumptions. The study uses fixed effects regression models to account for persistent neighborhood differences and broader citywide changes over time. The findings suggest that while simple regression models show a statistically significant relationship between arrests and overdose deaths, that relationship weakens substantially after accounting for neighborhood and time fixed effects. The results also suggest that overdose patterns and enforcement activity are not consistent across neighborhoods, indicating that broader local conditions and targeted public health interventions may play a larger role in explaining overdose risk 

2. Background & Related Literature 
The rise in overdose deaths across New York City has largely been driven by the increasing presence of fentanyl and other synthetic opioids within the drug supply. According to the New York City Department of Health and Mental Hygiene (2025), overdose deaths increased sharply beginning in the mid-2010s and accelerated further during the COVID-19 pandemic. Fentanyl became involved in overdose deaths throughout this period, contributing to record high mortality rates across many urban communities. The Department of Health and Mental Hygiene (2025) also notes that overdose deaths have disproportionately impacted neighborhoods facing higher levels of poverty, substance use, and limited access to healthcare resources. Recent research on overdose intervention strategies has also shifted attention toward more targeted public health approaches rather than traditional enforcement focused responses, particularly in neighborhoods experiencing persistent overdose hotspots.
Research on policing and drug enforcement has produced mixed conclusions regarding its effectiveness in reducing overdose deaths and drug activity. Supporters of enforcement based approaches argue that increased policing and drug arrests deter drug distribution and reduce the availability of dangerous substances in high-risk communities. These strategies are often connected to deterrence theory and “broken windows” style policing, which emphasize aggressive enforcement in areas experiencing high levels of crime and disorder. However, critics argue that arrest focused strategies fail to address the underlying causes of substance abuse, including poverty, homelessness, mental health conditions, and limited access to treatment. As a result, many researchers and policymakers have shifted toward public health approaches focused on treatment access, harm reduction, and targeted intervention programs.
Recent research has increasingly focused on prevention based and public health approaches to reducing overdose deaths rather than relying primarily on enforcement strategies. Many of these approaches emphasize early intervention, treatment access, harm reduction services, and targeted outreach within high-risk communities. Programs centered around overdose prevention and coordinated care have shown promising results in reducing both fatal and non-fatal overdoses, particularly in neighborhoods experiencing persistent overdose activity. This shift toward targeted intervention strategies reflects a broader understanding that overdose risk is often shaped by local social and economic conditions rather than policing levels alone. As a result, many policymakers have begun exploring more localized approaches aimed at identifying and assisting high-risk individuals before overdoses occur.
Given these competing explanations, this project uses regression analysis to examine the relationship between drug arrests and overdose deaths across New York City. The analysis begins with simple models before incorporating fixed effects and socioeconomic controls to account for broader neighborhood and citywide influences. To address suppressed overdose counts within the public health data, the study also tests multiple overdose assumptions to evaluate the consistency of the results across different specifications.

3. Data
This study combines overdose death data from the New York City Department of Health and Mental Hygiene, drug arrest data from the NYPD, and demographic and socioeconomic data from the American Community Survey between 2013 and 2021. The analysis focuses on overdose death rates and drug arrest rates across New York City neighborhoods while also incorporating controls for factors such as income, poverty, and unemployment. These variables were selected to account for broader economic and social conditions that may influence overdose risk and enforcement activity throughout the city.
Because some overdose counts were suppressed within the public health data to protect privacy in areas with low death counts, the analysis required additional assumptions when constructing overdose rate measures. To address this issue, the study estimates lower-bound, midpoint, and upper-bound overdose values for suppressed observations. This approach allows the analysis to test the sensitivity of the regression results under different assumptions regarding the missing data.


```{r}
library(tidyverse)
library(sf)
library(tigris)
library(readxl)
library(janitor)
library(lubridate)
library(stringr)

options(tigris_use_cache = TRUE)

```

Set Files Path

```{r}
arrest_path <- "C:/Users/jcas2/Downloads/NYPD_Arrests_Data_(Historic)_-_Dangerous_Drugs_20260331.csv"

acs1_path <- "C:/Users/jcas2/Downloads/jca100.github.io/acs1-data.csv"

acs5_path <- "C:/Users/jcas2/Downloads/jca100.github.io/acs5-data.csv"

od_file <- "C:/Users/jcas2/Downloads/jca100.github.io/arrests_cd_overdose_clean_2006_2021 (1).csv"

drug_path <- "C:/Users/jcas2/Downloads/jca100.github.io/nyc_cd_drug_related_2022_2023.csv"
```

# 2. READ DATA

```{r}
arrests <- read_csv(arrest_path, show_col_types = FALSE) |>
  mutate(
    ARREST_DATE = mdy(ARREST_DATE),
    arrest_year = year(ARREST_DATE)
  ) |>
  filter(!is.na(Longitude), !is.na(Latitude))

acs1 <- read_csv(acs1_path, show_col_types = FALSE) |>
  mutate(source = "acs1")

acs5 <- read_csv(acs5_path, show_col_types = FALSE) |>
  mutate(source = "acs5")

overdose_cd <- read_csv(od_file, show_col_types = FALSE)

drug_pop <- read_csv(drug_path, show_col_types = FALSE)
```

# 3. BUILD SPATIAL ARRESTS DATA

```{r}
arrests_sf <- arrests |>
  st_as_sf(
    coords = c("Longitude", "Latitude"),
    crs = 4326,
    remove = FALSE
  )

```

# 4. GEOGRAPHIES

```{r}
puma_sf_nyc <- tigris::pumas(state = "NY", year = 2020, cb = TRUE) |>
  st_transform(4326) |>
  rename(
    GEOID  = GEOID20,
    PUMACE = PUMACE20,
    NAME   = NAMELSAD20
  ) |>
  filter(str_detect(NAME, "Bronx|Brooklyn|Manhattan|Queens|Staten Island")) |>
  select(GEOID, PUMACE, NAME, geometry)

cd_sf <- st_read(
  "https://data.cityofnewyork.us/resource/5crt-au7u.geojson?$limit=1000",
  quiet = TRUE
) |>
  st_transform(4326) |>
  mutate(
    boro_cd = as.integer(boro_cd),
    boro_name = case_when(
      boro_cd %/% 100 == 1 ~ "Manhattan",
      boro_cd %/% 100 == 2 ~ "Bronx",
      boro_cd %/% 100 == 3 ~ "Brooklyn",
      boro_cd %/% 100 == 4 ~ "Queens",
      boro_cd %/% 100 == 5 ~ "Staten Island",
      TRUE ~ NA_character_
    )
  ) |>
  select(boro_cd, boro_name, geometry)

```

# 5. SPATIAL JOINS

```{r}

arrests_geo <- arrests_sf |>
  st_join(puma_sf_nyc, join = st_within, left = TRUE) |>
  st_join(cd_sf, join = st_within, left = TRUE)
```

# 6. CLEAN VARIABLES

```{r}
arrests_geo <- arrests_geo |>
  mutate(
    drug_category = case_when(
      PD_DESC == "CONTROLLED SUBSTANCE, POSSESSION 7" ~ "Controlled substance possession: 7th degree",
      PD_DESC == "CONTROLLED SUBSTANCE, POSSESSION 5" ~ "Controlled substance possession: 5th degree",
      PD_DESC == "CONTROLLED SUBSTANCE, POSSESSION 4" ~ "Controlled substance possession: 4th degree",
      PD_DESC == "CONTROLLED SUBSTANCE,POSSESS. 3" ~ "Controlled substance possession: 3rd degree",
      PD_DESC == "CONTROLLED SUBSTANCE,POSSESS. 2" ~ "Controlled substance possession: 2nd degree",
      PD_DESC == "CONTROLLED SUBSTANCE,POSSESS. 1" ~ "Controlled substance possession: 1st degree",
      PD_DESC %in% c(
        "CONTROLLED SUBSTANCE, POSSESSI",
        "CONTROLLED SUBSTANCE,POSSESS."
      ) ~ "Controlled substance possession: unspecified degree",
      PD_DESC == "DRUG, INJECTION OF" ~ "Drug injection offense",
      PD_DESC == "MARIJUANA, POSSESSION 4 & 5" ~ "Marijuana possession: lower degree",
      PD_DESC == "MARIJUANA, POSSESSION 1, 2 & 3" ~ "Marijuana possession: higher degree",
      PD_DESC == "USE CHILD TO COMMIT CONT SUB OFF" ~ "Aggravated drug offense involving child",
      TRUE ~ "Other or unknown drug offense"
    ),
    drug_category_broad = case_when(
      PD_DESC == "CONTROLLED SUBSTANCE, POSSESSION 7" ~ "controlled_substance_possession_low",
      PD_DESC %in% c(
        "CONTROLLED SUBSTANCE, POSSESSION 5",
        "CONTROLLED SUBSTANCE, POSSESSION 4"
      ) ~ "controlled_substance_possession_mid",
      PD_DESC %in% c(
        "CONTROLLED SUBSTANCE,POSSESS. 3",
        "CONTROLLED SUBSTANCE,POSSESS. 2",
        "CONTROLLED SUBSTANCE,POSSESS. 1"
      ) ~ "controlled_substance_possession_high",
      PD_DESC == "MARIJUANA, POSSESSION 4 & 5" ~ "marijuana_possession_low",
      PD_DESC == "MARIJUANA, POSSESSION 1, 2 & 3" ~ "marijuana_possession_high",
      PD_DESC %in% c(
        "DRUG, INJECTION OF",
        "USE CHILD TO COMMIT CONT SUB OFF",
        "CONTROLLED SUBSTANCE, POSSESSI",
        "CONTROLLED SUBSTANCE,POSSESS."
      ) ~ "other_drug_offense",
      TRUE ~ "other_or_unknown_drug_offense"
    ),
    law_cat_label = case_when(
      LAW_CAT_CD == "F" ~ "Felony",
      LAW_CAT_CD == "M" ~ "Misdemeanor",
      LAW_CAT_CD == "V" ~ "Violation",
      TRUE ~ "Other / missing"
    ),
    age_group_clean = case_when(
      AGE_GROUP %in% c("<18", "18-24", "25-44", "45-64", "65+") ~ AGE_GROUP,
      AGE_GROUP == "UNKNOWN" ~ "Unknown",
      TRUE ~ NA_character_
    ),
    perp_race_acs = case_when(
      PERP_RACE == "WHITE" ~ "white",
      PERP_RACE == "BLACK" ~ "black",
      PERP_RACE == "ASIAN / PACIFIC ISLANDER" ~ "asian",
      PERP_RACE %in% c("WHITE HISPANIC", "BLACK HISPANIC") ~ "hispanic",
      PERP_RACE %in% c("AMERICAN INDIAN/ALASKAN NATIVE", "OTHER", "UNKNOWN") ~ "other_or_unknown",
      TRUE ~ NA_character_
    ),
    perp_sex_label = case_when(
      PERP_SEX == "F" ~ "Female",
      PERP_SEX == "M" ~ "Male",
      PERP_SEX == "U" ~ "Unknown",
      TRUE ~ "Other / missing"
    )
  )

```

# 7. ACS PANEL AT PUMA-YEAR LEVEL

```{r}
acs_all <- bind_rows(
  acs1 |> filter(year <= 2008),
  acs5 |> filter(year >= 2009)
) |>
  mutate(
    GEOID = as.character(GEOID),
    year = as.integer(year)
  )

arrests_puma_agg <- arrests_geo |>
  st_drop_geometry() |>
  filter(!is.na(GEOID), !is.na(arrest_year)) |>
  group_by(GEOID, NAME, arrest_year) |>
  summarize(
    arrests_total = n(),
    arrests_felony = sum(law_cat_label == "Felony", na.rm = TRUE),
    arrests_misdemeanor = sum(law_cat_label == "Misdemeanor", na.rm = TRUE),
    cs_possession_low = sum(drug_category_broad == "controlled_substance_possession_low", na.rm = TRUE),
    cs_possession_mid = sum(drug_category_broad == "controlled_substance_possession_mid", na.rm = TRUE),
    cs_possession_high = sum(drug_category_broad == "controlled_substance_possession_high", na.rm = TRUE),
    mj_possession_low = sum(drug_category_broad == "marijuana_possession_low", na.rm = TRUE),
    mj_possession_high = sum(drug_category_broad == "marijuana_possession_high", na.rm = TRUE),
    other_drug_offense = sum(drug_category_broad == "other_drug_offense", na.rm = TRUE),
    arrests_female = sum(perp_sex_label == "Female", na.rm = TRUE),
    arrests_male = sum(perp_sex_label == "Male", na.rm = TRUE),
    arrests_under18 = sum(age_group_clean == "<18", na.rm = TRUE),
    arrests_18_24 = sum(age_group_clean == "18-24", na.rm = TRUE),
    arrests_25_44 = sum(age_group_clean == "25-44", na.rm = TRUE),
    arrests_45_64 = sum(age_group_clean == "45-64", na.rm = TRUE),
    arrests_65plus = sum(age_group_clean == "65+", na.rm = TRUE),
    arrests_white = sum(perp_race_acs == "white", na.rm = TRUE),
    arrests_black = sum(perp_race_acs == "black", na.rm = TRUE),
    arrests_asian = sum(perp_race_acs == "asian", na.rm = TRUE),
    arrests_hispanic = sum(perp_race_acs == "hispanic", na.rm = TRUE),
    .groups = "drop"
  )

arrests_acs <- arrests_puma_agg |>
  left_join(
    acs_all,
    by = c("GEOID", "arrest_year" = "year")
  )
```

# 8. COMMUNITY DISTRICT-YEAR PANEL

```{r}
arrests_cd_agg <- arrests_geo |>
  st_drop_geometry() |>
  filter(!is.na(boro_cd), !is.na(arrest_year)) |>
  group_by(boro_cd, boro_name, arrest_year) |>
  summarize(
    arrests_total = n(),
    arrests_felony = sum(law_cat_label == "Felony", na.rm = TRUE),
    arrests_misdemeanor = sum(law_cat_label == "Misdemeanor", na.rm = TRUE),
    cs_possession_low = sum(drug_category_broad == "controlled_substance_possession_low", na.rm = TRUE),
    cs_possession_mid = sum(drug_category_broad == "controlled_substance_possession_mid", na.rm = TRUE),
    cs_possession_high = sum(drug_category_broad == "controlled_substance_possession_high", na.rm = TRUE),
    mj_possession_low = sum(drug_category_broad == "marijuana_possession_low", na.rm = TRUE),
    mj_possession_high = sum(drug_category_broad == "marijuana_possession_high", na.rm = TRUE),
    other_drug_offense = sum(drug_category_broad == "other_drug_offense", na.rm = TRUE),
    arrests_female = sum(perp_sex_label == "Female", na.rm = TRUE),
    arrests_male = sum(perp_sex_label == "Male", na.rm = TRUE),
    arrests_under18 = sum(age_group_clean == "<18", na.rm = TRUE),
    arrests_18_24 = sum(age_group_clean == "18-24", na.rm = TRUE),
    arrests_25_44 = sum(age_group_clean == "25-44", na.rm = TRUE),
    arrests_45_64 = sum(age_group_clean == "45-64", na.rm = TRUE),
    arrests_65plus = sum(age_group_clean == "65+", na.rm = TRUE),
    arrests_white = sum(perp_race_acs == "white", na.rm = TRUE),
    arrests_black = sum(perp_race_acs == "black", na.rm = TRUE),
    arrests_asian = sum(perp_race_acs == "asian", na.rm = TRUE),
    arrests_hispanic = sum(perp_race_acs == "hispanic", na.rm = TRUE),
    .groups = "drop"
  )

overdose_cd_one <- overdose_cd |>
  mutate(
    boro_cd = as.integer(boro_cd),
    arrest_year = as.integer(arrest_year)
  ) |>
  group_by(boro_cd, arrest_year) |>
  summarize(
    overdose_total = sum(
      deaths_due_to_use_of_or_poisoning_by_psychoactive_substance_excluding_alcohol_and_tobacco,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

arrests_cd_overdose <- arrests_cd_agg |>
  left_join(overdose_cd_one, by = c("boro_cd", "arrest_year"))
```

# 9. LOOKUP TABLE FOR COMMUNITY DISTRICT NAMES

```{r}
cd_lookup <- tribble(
  ~boro_cd, ~cd_name,
  101, "Battery Park / Tribeca",
  102, "Greenwich Village / SoHo",
  103, "Lower East Side / Chinatown",
  104, "Chelsea / Clinton",
  105, "Midtown Business District",
  106, "Murray Hill / Stuyvesant Town",
  107, "Upper West Side",
  108, "Upper East Side",
  109, "Manhattanville / West Harlem",
  110, "Central Harlem",
  111, "East Harlem",
  112, "Washington Heights / Inwood",
  201, "Mott Haven / Melrose",
  202, "Hunts Point / Longwood",
  203, "Morrisania / Crotona",
  204, "Concourse / Highbridge",
  205, "Fordham / University Heights",
  206, "Belmont / East Tremont",
  207, "Kingsbridge Heights / Bedford",
  208, "Riverdale / Fieldston",
  209, "Parkchester / Soundview",
  210, "Throgs Neck / Co-op City",
  211, "Morris Park / Bronxdale",
  212, "Williamsbridge / Baychester",
  301, "Greenpoint / Williamsburg",
  302, "Fort Greene / Brooklyn Heights",
  303, "Bedford-Stuyvesant",
  304, "Bushwick",
  305, "East New York / Starrett City",
  306, "Park Slope / Carroll Gardens",
  307, "Sunset Park",
  308, "Crown Heights North / Prospect Heights",
  309, "Crown Heights South / Lefferts Gardens",
  310, "Bay Ridge / Dyker Heights",
  311, "Bensonhurst",
  312, "Borough Park",
  313, "Coney Island",
  314, "Flatbush / Midwood",
  315, "Sheepshead Bay",
  316, "Brownsville",
  317, "East Flatbush",
  318, "Flatlands / Canarsie",
  401, "Long Island City / Astoria",
  402, "Woodside / Sunnyside",
  403, "Jackson Heights",
  404, "Elmhurst / Corona",
  405, "Ridgewood / Maspeth",
  406, "Rego Park / Forest Hills",
  407, "Flushing",
  408, "Hillcrest / Fresh Meadows",
  409, "Kew Gardens / Woodhaven",
  410, "South Ozone Park / Howard Beach",
  411, "Bayside / Little Neck",
  412, "Jamaica / Hollis",
  413, "Queens Village",
  414, "Rockaway / Broad Channel",
  501, "St. George / Stapleton",
  502, "South Beach / Willowbrook",
  503, "Tottenville / Great Kills"
)

arrests_cd_overdose <- arrests_cd_overdose |>
  left_join(cd_lookup, by = "boro_cd") |>
  relocate(cd_name, .after = boro_cd)

```

# 10. MAP DATA

```{r}
cd_map <- cd_sf |>
  left_join(arrests_cd_overdose, by = "boro_cd")

cd_map_2021 <- cd_map |>
  filter(arrest_year == 2021)

```

# 11. SAVE CLEAN FILES

```{r}
write_csv(arrests_acs, "arrests_puma_acs_clean.csv")

arrests_cd_overdose_2006_2021 <- arrests_cd_overdose |>
  filter(arrest_year <= 2021)

write_csv(
  arrests_cd_overdose_2006_2021,
  "arrests_cd_overdose_clean_2006_2021.csv"
)
```

# 12. VISUALIZATIONS

# Overdose deaths in NYC over time

```{r}
city_year <- arrests_cd_overdose |>
  group_by(arrest_year) |>
  summarize(
    overdose_total = sum(overdose_total, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(city_year, aes(x = arrest_year, y = overdose_total)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Year",
    y = "Overdose Deaths",
    title = "Overdose Deaths in NYC Over Time"
  ) +
  theme_minimal()
```

# Drug arrests and overdoses over time

```{r}
city_year <- arrests_cd_overdose |>
  group_by(arrest_year) |>
  summarize(
    arrests_total = sum(arrests_total, na.rm = TRUE),
    overdose_total = sum(overdose_total, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(city_year, aes(x = arrest_year)) +
  geom_line(aes(y = arrests_total, color = "Arrests")) +
  geom_line(aes(y = overdose_total, color = "Overdoses")) +
  labs(
    x = "Year",
    y = "Count",
    color = "",
    title = "Drug Arrests and Overdose Deaths in NYC Over Time"
  ) +
  theme_minimal()
```

# Scatter plot

```{r}
ggplot(arrests_cd_overdose, aes(x = arrests_total, y = overdose_total)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  labs(
    title = "Higher Drug Arrests Are Associated with Lower Overdose Deaths",
    subtitle = "Community district-level relationship in NYC",
    x = "Drug-related arrests",
    y = "Overdose deaths"
  ) +
  theme_minimal()
```

# Borough comparison

```{r}
borough_2021 <- arrests_cd_overdose |>
  filter(arrest_year == 2021) |>
  group_by(boro_name) |>
  summarize(
    arrests_total = sum(arrests_total, na.rm = TRUE),
    overdose_total = sum(overdose_total, na.rm = TRUE),
    .groups = "drop"
  )

borough_long <- borough_2021 |>
  pivot_longer(
    cols = c(arrests_total, overdose_total),
    names_to = "type",
    values_to = "count"
  )

ggplot(
  borough_long,
  aes(x = reorder(boro_name, count), y = count, fill = type)
) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  labs(
    title = "Comparing Drug-Related Arrests and Overdose Deaths by Borough (2021)",
    subtitle = "Borough-level comparison in NYC",
    x = "",
    y = "Total Count",
    fill = ""
  ) +
  scale_fill_manual(
    values = c(
      "overdose_total" = "#F9844A",
      "arrests_total" = "#277DA1"
    ),
    labels = c("Overdose Deaths", "Drug-Related Arrests")
  ) +
  theme_minimal()
```

# Harlem Hot spot chart

```{r}
manhattan <- drug_pop |>
  filter(year == 2023, borough == "Manhattan") |>
  mutate(
    area = case_when(
      community_district %in% c(
        "Central Harlem",
        "East Harlem",
        "Manhattanville",
        "Manhattanville / West Harlem"
      ) ~ "Harlem",
      TRUE ~ "Other Manhattan"
    )
  )

ggplot(
  manhattan,
  aes(
    x = reorder(community_district, crude_rate_per_100k),
    y = crude_rate_per_100k,
    fill = area
  )
) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Harlem" = "#F94144",
      "Other Manhattan" = "grey70"
    )
  ) +
  labs(
    title = "Harlem Stands Out as an Overdose Hotspot",
    subtitle = "Crude overdose death rates per 100,000 residents in Manhattan (2023)",
    x = "",
    y = "Crude Rate per 100,000",
    fill = ""
  ) +
  theme_minimal()

```

Overdose deaths for Manhattan Community District

```{r}
manhattan <- drug_pop |>
  filter(year == 2023, borough == "Manhattan")

ggplot(
  manhattan,
  aes(
    x = reorder(community_district, crude_rate_per_100k),
    y = crude_rate_per_100k,
    fill = community_district
  )
) +
  geom_col(width = 0.7) +
  coord_flip() +
  labs(
    title = "Overdose Rates by Manhattan Community District (2023)",
    subtitle = "Crude overdose death rates per 100,000 residents",
    x = "",
    y = "Crude Rate per 100,000",
    fill = "Community District"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # removes messy legend

```

All NON-Manhattan boroughs our outside of Manhattan

```{r}
not_manhattan <- drug_pop |>
  filter(year == 2023, borough != "Manhattan")

ggplot(
  not_manhattan,
  aes(
    x = reorder(community_district, crude_rate_per_100k),
    y = crude_rate_per_100k,
    fill = borough
  )
) +
  geom_col(width = 0.7) +
  coord_flip() +
  labs(
    title = "Overdose Rates Outside Manhattan (2023)",
    subtitle = "Crude overdose death rates per 100,000 residents",
    x = "",
    y = "Crude Rate per 100,000",
    fill = "Borough"
  ) +
  theme_minimal()

```

All boroughs and neighborhoods

```{r}
all_nyc <- drug_pop |>
  filter(year == 2023)

ggplot(
  all_nyc,
  aes(
    x = reorder(community_district, crude_rate_per_100k),
    y = crude_rate_per_100k,
    fill = borough
  )
) +
  geom_col(width = 0.7) +
  coord_flip() +
  labs(
    title = "Overdose Rates by NYC Community District (2023)",
    subtitle = "Crude overdose death rates per 100,000 residents",
    x = "",
    y = "Crude Rate per 100,000",
    fill = "Borough"
  ) +
  theme_minimal()

```

# Overdose map

```{r}
ggplot(cd_map_2021) +
  geom_sf(aes(fill = overdose_total), color = "white", size = 0.2) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Overdose deaths"
  ) +
  labs(
    title = "Overdose Deaths by Community District (NYC, 2021)",
    caption = "Source: NYC DOHMH & NYPD data"
  ) +
  theme_minimal()

# Arrest map
ggplot(cd_map_2021) +
  geom_sf(aes(fill = arrests_total), color = "white", size = 0.2) +
  scale_fill_viridis_c(
    trans = "log",
    option = "viridis",
    name = "Drug arrests (log scale)"
  ) +
  labs(
    title = "Drug Arrests by Community District (NYC, 2021)",
    caption = "Source: NYPD Arrest Data"
  ) +
  theme_minimal()
```

# 13. LAG MODELS 

```{r}
analysis_lag <- arrests_cd_overdose |>
  arrange(boro_cd, arrest_year) |>
  group_by(boro_cd) |>
  mutate(
    arrests_lag1 = lag(arrests_total, 1),
    arrests_lag2 = lag(arrests_total, 2),
    arrests_lag3 = lag(arrests_total, 3)
  ) |>
  ungroup() |>
  mutate(
    log_overdose = log1p(overdose_total),
    log_arrests_lag1 = log1p(arrests_lag1),
    log_arrests_lag3 = log1p(arrests_lag3)
  )

model_lag1 <- lm(overdose_total ~ arrests_lag1, data = analysis_lag)
summary(model_lag1)

model_lag3 <- lm(overdose_total ~ arrests_lag3, data = analysis_lag)
summary(model_lag3)

model_log_lag1 <- lm(log_overdose ~ log_arrests_lag1, data = analysis_lag)
summary(model_log_lag1)

model_log_lag3 <- lm(log_overdose ~ log_arrests_lag3, data = analysis_lag)
summary(model_log_lag3)

model_control <- lm(
  log_overdose ~ log_arrests_lag1 + arrests_total,
  data = analysis_lag
)

summary(model_control)

ggplot(analysis_lag, aes(x = arrests_lag1, y = overdose_total)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Past Drug Arrests Are Associated with Lower Future Overdose Deaths",
    subtitle = "1-year lag relationship",
    x = "Drug arrests (previous year)",
    y = "Overdose deaths"
  ) +
  theme_minimal()
```
