---
title: "DANL 200: Introduction to Data Analytics<br>Project"
subtitle: "Quarto Template<br><br>" 
author: 
  - name: Jordan Alfano 
     
  - name: Jocelyn Matusiak 
   
  - name: Julia Candee 
  
  - name: Sophia Frisiras       
  - <br>
date: "`r Sys.Date()`"

format: 
  html:
    toc: true
    number-sections: true
    code-fold: false       # https://quarto.org/docs/output-formats/html-code.html
    code-summary: "Show the code"
    code-tools: true
    highlight-style: espresso # atom-one tango espresso

execute: 
  echo: true
  eval: true
  warning: false
  message: false
  fig-width: 9
  
---

```{r setup, include = FALSE}
library(tidyverse)
library(hrbrthemes)
library(skimr)

```


```{r}
day_wise <- read_csv('https://jem10126.github.io/day_wise.csv')
nvars <- format(round(ncol(day_wise), 0), 
                nsmall=0, 
                big.mark=",") 
nobs <- format(round(nrow(day_wise), 0), 
                nsmall=0, 
                big.mark=",")
```

- Here is the paged table using `rmarkdown::paged_table()` with the `results = 'asis'` chunk option.

```{r}
#| results: asis
#| echo: false
rmarkdown::paged_table(day_wise) 
```

<br>

The number of variables is `r nvars`; the number of observations is `r nobs`.

<br>


# Introduction


This project matters because it affected the rate at which people are contracting and healing from the COVID 19 virus at a daily rate. There is a greater amount of people who are recovering compared to the people who are dying. 


Our data visualization entails how both death and recovered are increasing at an exponential rate. This provides the assumption that while many may be dying there are more people recovering. In the real world this provides hope for those who are suffering from the virus.. 



## Data Visualization

```{r}
day_wise %>% 
  ggplot(aes(x = log(Deaths), 
             y = log(Recovered))) + 
  geom_point(alpha = .1, color = 'purple') +
  geom_smooth(method = lm, se = F) +
  theme_bw() +
  theme(legend.position = 'top')
```

```{r}
(skimr::skim(day_wise))
```

## Data Transformation

```{r}
day_wise <- day_wise %>% 
  group_by(Deaths, Recovered) %>% 
  summarise(Recovered_tot = sum(Recovered, na.rm = T),
            Deaths_mean = round(mean(Deaths, na.rm = T), 2),
            .groups = "drop")

```


```{r}
#| results: asis
#| echo: false
rmarkdown::paged_table(day_wise) 
```


<br><br><br><br>

# References

- [https://quarto.org/docs/output-formats/html-basics.html](https://quarto.org/docs/output-formats/html-basics.html)

- [https://quarto.org/docs/websites/](https://quarto.org/docs/websites/)

- [R for Data Science](http://r4ds.hadley.nz) by [Hadley Wickham](https://hadley.nz)

##Group Relevancy 
Sophia Frisiras helped on this project! She attented meetings and helped us whenever we needed some extra assitance! 


